# Tidying Data with tidyr {#tidying-data-with-tidyr}

## Introduction

Tidy data has one row per observation and one column per variable. Real data
rarely arrives that way: metrics sprawl across columns, several values share
one column, and combinations go missing. In this chapter we reshape data with
**tidyr**:

- `pivot_longer()` / `pivot_wider()` — move between long and wide layouts
- `separate_wider_delim()` / `unite()` — split one column into several and back
- `drop_na()` / `complete()` / `fill()` — handle missing values and combinations

We will use the following R packages:

```{r tidyr_libs, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
```

## Data

We reuse the ecommerce data from \@ref(dplyr-basics), so no new data to learn.
The URL and date examples borrow the `mock_strings` and `transact` data sets
already used in \@ref(strings-in-r) and \@ref(date-and-time-in-r).

```{r tidyr_data, message=FALSE}
ecom <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv',
    col_types = cols_only(device = col_factor(levels = c("laptop", "tablet", "mobile")),
      referrer = col_factor(levels = c("bing", "direct", "social", "yahoo", "google")),
      purchase = col_logical(), n_pages = col_double(), n_visit = col_double(),
      duration = col_double(), order_value = col_double(), order_items = col_double()
    )
  )

ecom
```

## Long format with `pivot_longer()`

Suppose we average three metrics by device. The result is *wide*: one column
per metric, which is awkward to plot or rank.

```{r tidyr_wide}
wide <- ecom |>
  summarise(across(c(n_pages, duration, order_value), \(x) mean(x, na.rm = TRUE)), .by = device)
wide
```

`pivot_longer()` stacks the metric columns into `metric`/`mean_value` pairs —
one row per device-metric observation:

```{r tidyr_long}
long <- wide |>
  pivot_longer(-device, names_to = "metric", values_to = "mean_value")
long
```

## Wide format with `pivot_wider()`

The reverse trip is just as common. Counts by device and purchase come out
long (one row per combination):

```{r tidyr_counts}
counts <- ecom |>
  count(device, purchase)
counts
```

`pivot_wider()` spreads `purchase` into columns so each device occupies a
single row — the layout managers usually want in a report:

```{r tidyr_wider}
counts |>
  pivot_wider(names_from = purchase, values_from = n, names_prefix = "purchase_")
```

## Split columns with `separate_wider_delim()`

The `url` column of `mock_strings` packs the protocol and the address into one
string, separated by `://`. `separate_wider_delim()` splits it into two
columns:

```{r tidyr_separate_url, message=FALSE}
mockstring <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/mock_strings.csv')

tibble(url = head(mockstring$url, 3)) |>
  separate_wider_delim(url, delim = "://", names = c("protocol", "rest")) |>
  select(protocol)
```

The same verb splits calendar dates. The `Due` column of `transact` is
`YYYY-MM-DD`; splitting on `"-"` recovers the components:

```{r tidyr_separate_date, message=FALSE}
transact <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/transact.csv')

tibble(due_iso = as.character(head(transact$Due, 5))) |>
  separate_wider_delim(due_iso, delim = "-", names = c("due_year", "due_month", "due_day"))
```

## Combine columns with `unite()`

`unite()` is the inverse: it pastes several columns back into one. Splitting
the ISO date and reuniting it returns the input exactly — a useful check that
a separate/unite round-trip loses nothing:

```{r tidyr_unite}
tibble(due_iso = as.character(head(transact$Due, 5))) |>
  separate_wider_delim(due_iso, delim = "-", names = c("due_year", "due_month", "due_day")) |>
  unite("due_rebuilt", due_year, due_month, due_day, sep = "-")
```

## Drop missing values with `drop_na()`

`ecom` itself has no missing values, but joins create them: the full join of
`customer` and `order` from \@ref(joining-tables-in-r-dplyr) carries `NA`s for
every unmatched row. `drop_na()` keeps only complete rows:

```{r tidyr_dropna, message=FALSE}
order <- read_delim('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/order.csv', delim = ';')
customer <- read_delim('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/customer.csv', delim = ';')

joined <- full_join(customer, order, by = join_by(id))
nrow(joined)

nrow(drop_na(joined))
```

Only the 55 fully matched rows survive — the same rows `inner_join()` returns.
Use `drop_na()` when a downstream step needs complete cases, and prefer it over
filtering each column by hand.

## Complete combinations with `complete()`

Counts over a filtered subset can silently omit combinations. Purchasers only
span `purchase == TRUE`, so the `FALSE` rows are missing rather than zero:

```{r tidyr_complete}
purchasers <- ecom |>
  filter(purchase) |>
  count(device, purchase)
purchasers
```

`complete()` adds the missing combinations and fills the count with 0, so
plots and tables show every device × purchase cell:

```{r tidyr_complete_fill}
purchasers |>
  complete(device, purchase = c(TRUE, FALSE), fill = list(n = 0L))
```

## Fill down labels with `fill()`

Spreadsheet exports often write a group label once and leave the rest blank.
`fill()` carries the last non-missing value down (or up with
`.direction = "up"`):

```{r tidyr_fill}
report <- tibble(device = c("laptop", NA, "tablet", NA),
                 metric = c("aov", "orders", "aov", "orders"))
report |>
  fill(device)
```

## Your Turn

- reshape the mean of `n_visit` and `order_items` by `referrer` from wide to long
- split the `email` column of `mockstring` at `"@"` and reunite it — does the round-trip hold?
- run the `complete()` example without `fill`: what does the missing count become, and why?
