# Arrow, Parquet and DuckDB {#arrow-duckdb}

## Introduction

Everything so far fit in memory as CSV. Two tools cover the next scale, and
both speak the dplyr you already know:

- **Parquet** (via **arrow**) — a compressed columnar file format. The same
data takes far less disk and reads much faster than CSV.
- **DuckDB** (via **duckplyr**) — an in-process analytical database. dplyr
verbs run inside it, so data larger than memory stays on disk until
`collect()`.

Both are optional: install them with
`install.packages(c("arrow", "duckdb", "duckplyr"))`. The chunks below run only
when the packages are present (`eval = requireNamespace(...)`), so the book
still builds without them. Browser execution of parquet is intentionally
deferred until WebR binary sizes are checked (see the P2 plan).

## Parquet with arrow

Write `ecom` to parquet and read it back. The round-trip lands in a temporary
directory, so nothing litters the project:

```{r arrow_parquet, message=FALSE, eval=requireNamespace("arrow", quietly=TRUE)}
library(dplyr)
library(readr)
library(arrow)

ecom <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv',
    col_types = cols_only(device = col_factor(levels = c("laptop", "tablet", "mobile")),
      referrer = col_factor(levels = c("bing", "direct", "social", "yahoo", "google")),
      purchase = col_logical(), n_pages = col_double(), n_visit = col_double(),
      duration = col_double(), order_value = col_double(), order_items = col_double()
    )
  )

pq <- file.path(tempdir(), "ecom.parquet")
cs <- file.path(tempdir(), "ecom.csv")
write_parquet(ecom, pq)
write_csv(ecom, cs)

# same rows, roughly half the bytes
c(csv_bytes = file.size(cs), parquet_bytes = file.size(pq))
```

`read_parquet()` returns a tibble, so every verb from
\@ref(dplyr-basics) works unchanged:

```{r arrow_read, eval=requireNamespace("arrow", quietly=TRUE)}
ecom_pq <- read_parquet(pq)
class(ecom_pq)

ecom_pq |>
  filter(purchase) |>
  summarise(revenue = sum(order_value), orders = n(), .by = device)
```

## DuckDB with duckplyr

`duckdb_tibble()` (duckplyr >= 1.0) moves a data frame into DuckDB and hands
back a lazy table: dplyr verbs build the query, nothing runs until
`collect()`:

```{r duckdb_aov, message=FALSE, eval=requireNamespace("duckplyr", quietly=TRUE)}
library(duckplyr)

ecom_duck <- duckdb_tibble(ecom)
class(ecom_duck)

ecom_duck |>
  filter(purchase) |>
  summarise(revenue = sum(order_value), orders = n(), .by = device) |>
  mutate(aov = revenue / orders) |>
  collect()
```

The numbers match the Chapter \@ref(dplyr-basics) AOV exactly — the engine
changed, the answers did not. Reach for this pattern when a single CSV no
longer fits comfortably in memory: keep the data in DuckDB (or parquet on
disk), push the verbs down, and `collect()` only the summary.
