# Exercise Solutions {#exercise-solutions}

## Introduction

Try every **Your Turn** before reading its solution — the struggle is the
lesson. Worked solutions below are seeded for \@ref(dplyr-basics); solutions
for the remaining chapters are open stubs. To contribute one, open an issue or
pull request with the chapter, the question, and a runnable chunk in this
book's style (`|>`-first, same data as the chapter).

We will use the following R packages:

```{r sol_libs, message=FALSE}
library(dplyr)
library(readr)
```

## dplyr Basics

```{r sol_data, message=FALSE}
ecom <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv',
    col_types = cols_only(device = col_factor(levels = c("laptop", "tablet", "mobile")),
      referrer = col_factor(levels = c("bing", "direct", "social", "yahoo", "google")),
      purchase = col_logical(), n_pages = col_double(), n_visit = col_double(),
      duration = col_double(), order_value = col_double(), order_items = col_double()
    )
  )
```

**Q1. What is the average number of pages visited by purchasers and
non-purchasers?**

```{r sol_dplyr1}
ecom |>
  summarise(mean_pages = mean(n_pages), .by = purchase)
```

Purchasers browse far more pages (~16 vs ~5) — the signal behind the AOV case
study.

**Q2. What is the average time on site for purchasers vs non-purchasers?**

```{r sol_dplyr2}
ecom |>
  summarise(mean_time = mean(duration), .by = purchase)
```

Time on site barely differs (~355 vs ~359 seconds): page depth, not duration,
separates buyers here.

**Q3. What is the average number of pages visited by purchasers and
non-purchasers using mobile?**

```{r sol_dplyr3}
ecom |>
  filter(device == "mobile") |>
  summarise(mean_pages = mean(n_pages), .by = purchase)
```

## Open stubs

Solutions for these chapters are not yet written — contributions welcome
(see above):

- \@ref(joining-tables-in-r-dplyr): no **Your Turn** block; propose join puzzles on `customer`/`order`
- \@ref(dplyr-helper-functions): sampling/slicing drills on `ecom`
- \@ref(r-pipe-magrittr): rewrite-a-chain exercises (`%>%` → `|>`, `_`, `\(x)`)
- \@ref(tibbles-in-r): tibble vs data.frame edge cases
- \@ref(tidying-data-with-tidyr): the three **Your Turn** prompts in that chapter
- \@ref(strings-in-r): URL/email parsing variants on `mockstring`
- \@ref(date-and-time-in-r): the nine **Your Turn** blocks (intervals, TZ/DST, formats)
- \@ref(categorical-data-in-r): the seven **Your Turn** blocks (lumping, reordering, recoding)
