---
description: "Data Wrangling in R: worked solutions to chapter exercises, plus open stubs for contributors."
---

# Exercise Solutions {#exercise-solutions}

## Introduction

Try every **Your Turn** before reading its solution — the struggle is the
lesson. Worked solutions below cover @dplyr-basics, @tidying-data-with-tidyr,
@date-and-time-in-r, @date-arithmetic-time-zones, @categorical-data-in-r and
@forcats-manipulation. Remaining chapters are open stubs — to contribute one,
open an issue or pull request with the chapter, the question, and a runnable
chunk in this book's style (`|>`-first, same data as the chapter).

We will use the following R packages:

```{r sol_libs, message=FALSE}
library(dplyr)
library(readr)
```

## dplyr Basics

```{r sol_data, message=FALSE}
ecom <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv',
    col_types = cols_only(device = col_factor(levels = c("laptop", "tablet", "mobile")),
      referrer = col_factor(levels = c("bing", "direct", "social", "yahoo", "google")),
      purchase = col_logical(), n_pages = col_double(), n_visit = col_double(),
      duration = col_double(), order_value = col_double(), order_items = col_double()
    )
  )
```

**Q1. What is the average number of pages visited by purchasers and
non-purchasers?**

```{r sol_dplyr1}
ecom |>
  summarise(mean_pages = mean(n_pages), .by = purchase)
```

Purchasers browse far more pages (~16 vs ~5) — the signal behind the AOV case
study.

**Q2. What is the average time on site for purchasers vs non-purchasers?**

```{r sol_dplyr2}
ecom |>
  summarise(mean_time = mean(duration), .by = purchase)
```

Time on site barely differs (~355 vs ~359 seconds): page depth, not duration,
separates buyers here.

**Q3. What is the average number of pages visited by purchasers and
non-purchasers using mobile?**

```{r sol_dplyr3}
ecom |>
  filter(device == "mobile") |>
  summarise(mean_pages = mean(n_pages), .by = purchase)
```

## Tidying Data with tidyr

```{r sol_tidyr_data, message=FALSE}
library(tidyr)
mockstring <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/mock_strings.csv',
    show_col_types = FALSE)
```

**Q1. Reshape the mean of `n_visit` and `order_items` by `referrer` from wide
to long.**

```{r sol_tidyr1}
ecom |>
  summarise(across(c(n_visit, order_items), \(x) mean(x, na.rm = TRUE)),
    .by = referrer) |>
  pivot_longer(-referrer, names_to = "metric", values_to = "mean_value")
```

**Q2. Split the `email` column of `mockstring` at `"@"` and reunite it — does
the round-trip hold?**

```{r sol_tidyr2}
split <- mockstring |>
  select(email) |>
  separate(email, into = c("user", "domain"), sep = "@", remove = FALSE)
reunited <- split |>
  unite("rejoined", user, domain, sep = "@", remove = TRUE)
identical(reunited$rejoined, mockstring$email)
```

Yes — every row round-trips (`TRUE`), since each address holds exactly one
`"@"`.

**Q3. Run the `complete()` example without `fill`: what does the missing count
become, and why?**

```{r sol_tidyr3}
ecom |>
  filter(purchase) |>
  count(device, purchase) |>
  complete(device, purchase = c(TRUE, FALSE))
```

The missing cells become `NA`, not 0: with no `fill`, `complete()` adds the
absent combinations but has no value to put in `n`, so the count is unknown
(`NA`). Pass `fill = list(n = 0L)` when a missing combination genuinely means
"zero events".

## Date & Time: Parsing and Components

```{r sol_lub10_libs, message=FALSE}
library(lubridate)
transact <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/transact.csv',
    show_col_types = FALSE)
```

**Q1. Current date, current time, AM or PM, and leap years.**

```{r sol_lub10a}
today()
now()
pm(now())
leap_year(c(2018, 2016))
```

2018 was not a leap year (`FALSE`); 2016 was (`TRUE`).

**Q2. R 1.0.0 was released on `2000-02-29 08:55:23 UTC`. Save it as a `Date`
from character, a `Date` from origin + number, `POSIXct`, `POSIXlt`
(extracting month day, day of year, month, zone) and `ISODate`.**

```{r sol_lub10b}
as.Date("2000-02-29")
as.Date(11016, origin = "1970-01-01")
as.POSIXct("2000-02-29 08:55:23", tz = "UTC")
lt <- as.POSIXlt("2000-02-29 08:55:23", tz = "UTC")
c(month_day = lt$mday, day_of_year = lt$yday, month = lt$mon + 1L, zone = lt$zone)
ISOdate(2000, 2, 29, 8, 55, 23, tz = "UTC")
```

`$mon` is zero-indexed, hence `+ 1L`; `$yday` is 59 (zero-indexed — the 60th
day of the leap year).

**Q3. Create `July 5th, 2019` from each spelling with `as.Date()` and the
correct format.**

```{r sol_lub10c}
as.Date("July-05-19", format = "%B-%d-%y")
as.Date("JUL-05-19", format = "%b-%d-%y")
as.Date("05.07.19", format = "%d.%m.%y")
as.Date("5-July 2019", format = "%d-%B %Y")
as.Date("July 5th, 2019", format = "July %dth, %Y")
as.Date("July 05, 2019", format = "%B %d, %Y")
as.Date("2019-July- 05", format = "%Y-%B- %d")
as.Date("05/07/2019", format = "%d/%m/%Y")
as.Date("07/05/2019", format = "%m/%d/%Y")
as.Date("7/5/2019", format = "%m/%d/%Y")
as.Date("07/5/19", format = "%m/%d/%y")
as.Date("2019-07-05")
```

Note the ambiguous pairs: `05/07/2019` vs `07/05/2019` is day-first vs
month-first — the format string, not the data, decides. `as.Date()` matches
literals (`"July "`, `"th"`, `", "`) character-for-character.

**Q4. Parse the same spellings with `strptime()` / `parse_date_time()` /
helpers.**

```{r sol_lub10d}
dates <- c("July-05-19", "JUL-05-19", "05.07.19", "5-July 2019",
  "July 5th, 2019", "July 05, 2019", "2019-July- 05", "05/07/2019",
  "07/05/2019", "7/5/2019", "07/5/19", "2019-07-05")
parse_date_time(dates, orders = c("Bdy", "bdy", "dmy", "dBY", "BYd",
  "dmy", "mdy", "mdy", "mdy", "mdy", "ymd"))
mdy("July 05, 2019")
dmy("05.07.19")
ymd("2019-07-05")
```

One `parse_date_time()` call with several `orders` handles the whole mixed
vector — prefer helpers (`mdy()`/`dmy()`/`ymd()`) when a column uses a single
convention.

**Q5. From the `Due` column of `transact`, tabulate year, labelled month,
labelled weekday, hour and quarter.**

```{r sol_lub10e}
transact |>
  mutate(due_year = year(Due),
    due_month = month(Due, label = TRUE),
    due_wday = wday(Due, label = TRUE),
    due_hour = hour(Due),
    due_quarter = quarter(Due)) |>
  count(due_year, due_month, due_wday, due_hour, due_quarter)
```

`due_hour` is 0 everywhere: `Due` is a `Date` (midnight, no time component) —
`hour()` on dates is always 0.

**Q6. R 2.0.0 (`2004-10-04 14:24:38`) with `make_date()` and `make_datetime()`;
R 3.0.0 (`2013-04-03 07:12:36`) via `update()`.**

```{r sol_lub10f}
r200 <- make_datetime(2004, 10, 4, 14, 24, 38)
r200_date <- make_date(2004, 10, 4)
update(r200, year = 2013, month = 4, mday = 3, hour = 7, min = 12, sec = 36)
```

## Date & Time: Arithmetic and Time Zones

**Q1. Vacation length, shifted vacation, invoice settlement and employment
length.**

```{r sol_lub11a, message=FALSE}
vac_start <- ymd("2020-04-19")
vac_end <- ymd("2020-04-25")
vac_start %--% vac_end
as.period(vac_start %--% vac_end)

# shift the start by 10 days and the end by 2 weeks, then recompute
vac_start2 <- vac_start + days(10)
vac_end2 <- vac_end + weeks(2)
as.period(vac_start2 %--% vac_end2)

receivables <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/receivables.csv',
    show_col_types = FALSE)
receivables |>
  mutate(invoice_date = mdy(invoice_date), due_date = mdy(due_date),
    settled_date = mdy(settled_date),
    days_to_settle = settled_date - invoice_date,
    days_overdue = settled_date - due_date) |>
  select(invoice_number, days_to_settle, days_overdue) |>
  head()

hr <-
  read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/hr-data.csv',
    show_col_types = FALSE)
hr |>
  filter(Termd == 1, !is.na(DateofTermination)) |>
  mutate(hire = mdy(DateofHire), term = mdy(DateofTermination),
    employed = term - hire) |>
  select(Employee_Name, hire, term, employed) |>
  head()
```

The base vacation is 6 days; after shifting it spans 10 days. Positive
`days_overdue` means the customer paid late; filter `Termd == 1` (plus a
non-missing termination date) so only terminated employees count.

**Q2. Your timezone, daylight savings and the current UTC time.**

```{r sol_lub11b}
Sys.timezone()
dst(now())
now("UTC")
```

**Q3. Round `transact` due dates: up to weeks, down to days, back to the start
of the month.**

```{r sol_lub11c}
transact |>
  mutate(due_week_up = ceiling_date(Due, "week"),
    due_day_down = floor_date(Due, "day"),
    due_month_start = rollback(Due, roll_to_first = TRUE)) |>
  select(Due, due_week_up, due_day_down, due_month_start) |>
  head()
```

`Due` values are whole dates, so rounding down to days is a no-op (midnight
already); `ceiling_date()` moves each due date forward to its week boundary.

## Categorical Data Fundamentals

```{r sol_fct12_data, message=FALSE}
analytics_raw <- readRDS("analytics_raw.rds")
```

**Q1. Quantitative or qualitative?** Quantitative: Age, Annual Sales, Weight,
Height, Job/Work Experience. Qualitative: Gender, Education Level, Vehicle
Type, Ice Cream Flavor, Blood Group, Color, Mode of Payment. (Education Level
is qualitative but *ordered* — see ordered factors below.)

**Q2. Dichotomous or polychotomous?** Dichotomous (exactly two outcomes):
Heads / Tails, Rich / Poor, Democrat / Republican. Polychotomous (three or
more): Blood Group, Transportation Mode, Education Level.

**Q3–Q9. Factor checks, coercion and level control on `analytics_raw`.**

```{r sol_fct12a}
# Q3: which of these are already factors? (raw file: none — all character)
sapply(analytics_raw[c("device", "page_depth", "landing_page")], is.factor)

# Q4: coerce the categorical columns (page_depth stays numeric;
# `city` is not in this file, so it is skipped)
cat_vars <- c("device", "os", "browser", "user_type", "channel",
  "gender", "landing_page", "exit_page", "country")
analytics_raw[cat_vars] <- lapply(analytics_raw[cat_vars], factor)

# Q5: keep only male/female levels in gender
gender <- factor(analytics_raw$gender, levels = c("male", "female"))
levels(gender)

# Q6: admit NA as a level (no-op here — this file has no missing gender,
# but the call is what makes unseen NAs explicit downstream)
library(forcats)
gender <- fct_na_value_to_level(gender, level = NA)
levels(gender)

# Q7: label that NA level "missing"
gender <- fct_na_value_to_level(gender, level = "(Missing)")
levels(gender)

# Q8: relabel user_type levels
user_type <- fct_recode(factor(analytics_raw$user_type),
  "New" = "New Visitor", "Returning" = "Returning Visitor")
levels(user_type)

# Q9: is user_rating ordered? coerce if not
is.ordered(analytics_raw$user_rating)
user_rating <- ordered(analytics_raw$user_rating)
is.ordered(user_rating)
```

**Q10–Q13. Level counts, categories, frequencies and a two-way summary.**

```{r sol_fct12b}
# Q10: number of levels
sapply(analytics_raw[c("browser", "channel", "landing_page", "exit_page")],
  \(x) nlevels(factor(x)))

# Q11: the categories themselves
lapply(analytics_raw[c("os", "channel", "browser", "gender", "user_type")],
  \(x) levels(factor(x)))

# Q12: frequencies
fct_count(factor(analytics_raw$channel))
fct_count(factor(analytics_raw$user_type))

# Q13: channel by user_type — counts and column shares
table(factor(analytics_raw$channel), factor(analytics_raw$user_type))
proportions(table(factor(analytics_raw$channel),
  factor(analytics_raw$user_type)), margin = 2)
```

## Factor Manipulation with forcats

```{r sol_fc13_data, message=FALSE}
library(forcats)
data <- readRDS("analytics.rds")
```

**Q1–Q2. Frequencies in descending order; is `laptop` a `device` level?**

```{r sol_fc13a}
data |> pull(device) |> fct_count()
data |> pull(landing_page) |> fct_count()
data |> pull(exit_page) |> fct_count()
"laptop" %in% levels(data$device)
```

`fct_count()` sorts most-frequent-first. (`device` here holds
Desktop/Mobile/Tablet, so `"laptop"` is not a level — it belongs to the
`ecom` schema, not `analytics`.)

**Q3–Q8. Collapsing, lumping, keeping and anonymizing levels.**

```{r sol_fc13b}
# Q3: collapse account-related pages
data |> pull(landing_page) |>
  fct_collapse(Account = c("My Account", "Register", "Sign In", "Your Info")) |>
  fct_count() |> head(6)

# Q4: lump landing pages behind fewer than 1000 visits into "Other"
data |> pull(landing_page) |> fct_lump_min(min = 1000) |> fct_count()

# Q5: top 10 landing and exit pages
data |> pull(landing_page) |> fct_lump_n(n = 10) |> fct_count()
data |> pull(exit_page) |> fct_lump_n(n = 10) |> fct_count()

# Q6: landing pages with at least 5% of traffic
data |> pull(landing_page) |> fct_lump_prop(prop = 0.05) |> fct_count()

# Q7: keep only the big-four browsers
data |> pull(browser) |>
  fct_other(keep = c("Chrome", "Firefox", "Safari", "Edge")) |>
  fct_count()

# Q8: anonymize page levels (stable within a session via seed)
set.seed(42)
data |> pull(landing_page) |> fct_anon() |> head()
```

**Q9–Q15. Reordering, shuffling and reversing levels.**

```{r sol_fc13c}
set.seed(42)
landing_page <- data$landing_page
browser <- data$browser

# Q9/Q10/Q11: Home first, Apparel second, Specials last
landing_page |> fct_relevel("Home") |> levels() |> head(4)
landing_page |> fct_relevel("Apparel", after = 1L) |> levels() |> head(4)
landing_page |> fct_relevel("Specials", after = Inf) |> levels() |> tail(3)

# Q12: browsers by frequency
browser |> fct_infreq() |> levels()

# Q13: landing pages in order of appearance
landing_page |> fct_inorder() |> levels() |> head(6)

# Q14: shuffled os levels
data |> pull(os) |> fct_shuffle() |> levels()

# Q15: reversed browser levels
browser |> fct_rev() |> levels()
```

**Q16. The plots.**

```{r sol_fc13d1, fig.height=5, fig.width=7}
library(ggplot2)
# 1: bar plot of channel
ggplot(data) +
  geom_bar(aes(x = channel), fill = "blue") +
  ggtitle("Source of Traffic") + xlab("Channel") + ylab("Count")
# 2: grouped bars, user_type by channel
ggplot(data) +
  geom_bar(aes(x = channel, fill = user_type), position = "dodge") +
  ggtitle("User Type by Channel") + xlab("Channel") + ylab("User Type")
```

```{r sol_fc13d2, fig.height=5, fig.width=7}
# 3: stacked bars, channel by gender
ggplot(data) +
  geom_bar(aes(x = gender, fill = channel)) +
  ggtitle("Channel by Gender") + xlab("Gender") + ylab("Channel")
# 4: proportional bars, channel by device
data |>
  select(device, channel) |>
  table() |>
  tibble::as_tibble() |>
  ggplot(aes(x = device, y = n, fill = channel)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Channel by Device") + xlab("Device") + ylab("Channel")
```

```{r sol_fc13d3, fig.height=5, fig.width=7}
# 5: mosaic plots need ggmosaic (archived on CRAN) — out of scope;
# see the ggplot2 ebook instead.
# 6: pie and donut of channel
data |> pull(channel) |> table() |> pie()
data |>
  pull(channel) |>
  fct_count() |>
  rename(channel = f, count = n) |>
  ggplot() +
  geom_bar(aes(x = "", y = count, fill = channel), width = 1,
    stat = "identity") +
  coord_polar("y", start = 0)
```

## Open stubs

Chapters without **Your Turn** blocks still need proposed drills + solutions —
contributions welcome (see above):

- @joining-tables-in-r-dplyr: join puzzles on `customer`/`order`
- @dplyr-helper-functions: sampling/slicing drills on `ecom`
- @r-pipe-magrittr: rewrite-a-chain exercises (`%>%` → `|>`, `_`, `\(x)`)
- @tibbles-in-r: tibble vs data.frame edge cases
- @strings-in-r: URL/email parsing variants on `mockstring`
- @import-data-in-r-basics and @import-data-in-r-advanced: import-parameter drills
